---
title: "Data Cleaning"
author: "Daniel"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(stringr)
```

### Dataset Loading

```{r}
df <- read.csv("raw_data.csv")
head(df)
```

### Sorting the Dataset

I will sort the records by their ID.

```{r}
df <- df |> arrange(ID)
```
 
### Loading "Clean" Names

After asking the AI to clean the names with symbols, I load the processed name data.

```{r}
df$name <- read.csv("cleaned_names.csv")$name
```

### Cleaning the `name` column

```{r}
df$name <- str_to_title( str_trim(df$name) )
```

### Converting `gender` values

```{r}
df$gender <- as.factor(df$gender) |> recode("M" = "Male", "F" = "Female")
```

### Converting Department to Uppercase

```{r}
df$department <- str_to_upper(df$department)
```

### Converting Salary to Decimal Numbers

```{r}
df$salary <- str_replace(df$salary, "\$", "") |> as.numeric()
```

We can verify that although RStudio represents numbers as integers, they are actually decimal numbers.

```{r}
sum(round(df$salary) != df$salary, na.rm=T)
```

### Converting Dates

```{r}
df$contract_start_date <- parse_date_time(df$contract_start_date,
                orders = c("m/d/Y", "Y-m-d HMS")) |> 
  ymd()
```

### Recoding `tenure` and `contract_type`

#### `tenure`:
  - True: Full-time job
  - False: Part-time job
  
#### `contract_type`:
  - True: Permanent
  - False: Fixed-term

```{r}
df$tenure <- ifelse(df$tenure, "full time", "part time")
df$contract_type <- ifelse(df$contract_type, "permanent", "fixed-term")
```

### Cleaning Country of Origin

Before starting, let's take a look at the unique values for the `country` column:

```{r}
unique(df$country)
```

It is necessary to clean extra spaces and remove the `?` character from some values.

```{r}
df$country <- str_replace_all(df$country, "\?", "") |> # remove ?
  str_trim() |>  # remove extra spaces
  str_to_title() |>  # convert to title case
  as.factor() |>  # convert to factor (to recode values later)
  fct_recode(  # recoding some misspelled values
    "Spain"   = "Spian",
    "Spain"   = "Sapin",
    "France"  = "Fracne",
    "France"  = "Frnace",
    "France"  = "Fra",
    "China"   = "Chian",
    "China"   = "Chnia",
    "Italy"   = "Italia",
    "Italy"   = "Tialy",
    "Italy"   = "It",
    "Japan"   = "Japán",
    "Japan"   = "Jpn",
    "Brazil"  = "Brasil",
    "USA"     = "Usa"
  )
```

### Email Address Creation

Using the employee's name initials and their ID, I will construct the corresponding email addresses.

The first step will be to separate each name into columns.

```{r}
separate_wider_delim(df |> head(), cols = "name", " ", names_sep = "_") |> 
  select(starts_with("name"))
```

Once we have verified that the method works, we apply it to the entire dataframe.

```{r}
# generates an error!
# separate_wider_delim(df, cols = "name", " ", names_sep = "_")
```

An error occurs because some names have 5 words. I will inspect those names...

```{r}
splitted_names <- str_split(df$name, " ")
splitted_names |> head()
```

Searching for those with more than 3 words:

```{r}
splitted_names[lengths(splitted_names) > 3]
```

I will obtain the indexes for those names:

```{r}
(indexes <- which (lengths(splitted_names) > 3))
```

The problematic Spanish names do not make sense from a linguistic perspective. I will proceed to modify them manually.

```{r}
shortened_names <- c("Francisco José Fernández",
                     "María Dolores Garcia",
                     "María Dolores López",
                     "María Carmen Gómez",
                     "Ana María Sánchez",
                     "José Luis Fernández",
                     "José Luis Rodríguez",
                     "Teresa María Martínez",
                     "José Luis Fernández",
                     "Josefa María Pérez",
                     "Yan Yan Zhag",
                     "José Antônio Lima",
                     "Laura Maria Rodríguez",
                     "María Carmen Pérez")

# create a dataframe with the indexes and corrected values
(shortened_names_df <- data.frame("index" = indexes,
                                  "original_name" = df[indexes, "name"], 
                                 "name" = shortened_names) )
```
I will export these shortened names into a csv file. It will be handy for the cleaning process using Excel.

```{r}
shortened_names_df |> write.csv("shortened_names.csv", row.names = F)
```


Now I will substitute them in the `name` column:

```{r}
df[shortened_names_df$index, "name"] <- shortened_names_df$name
```

Attempting to separate name components again:

```{r}
separate_wider_delim(df, cols = "name", " ", names_sep = "_") |> 
  select(starts_with("name"))
```

Now everything is correct. I want to obtain the initial of each name component.

```{r}
df <- separate_wider_delim(df, cols = "name", delim= " ",
                           names_sep = "_", cols_remove = F) |> 
  # reorganize columns
  rename( "name" = "name_name") |> 
  relocate(name, .before = "name_1")

df |> head()
```

I want to obtain the initial of each name component.

```{r}
df <- df |>
  # use columns starting with name_
  # and extract the initial from each one
  mutate(across(starts_with("name_"), ~ str_sub(., 1,1))) |> 
  # join these columns using a separator: ""
  # columns name_1, name_2 and name_3 will disappear after the join
  unite("initials", starts_with("name_" ), sep="" ) |>
  # convert them to lowercase
  mutate(initials = str_to_lower(initials) )

df |> head()
```

Finally, we construct the email: `initials.ID@clean_data.com`

```{r}
df <- df |> mutate( email = paste0(initials, ".", ID, "@clean_data.com") )

# remove the initials column
df <- df |> select(-initials)

# reorder columns
df <- df |>
  relocate(email, .before="name") |> 
  relocate(country, .after="name")
```

With the clean dataset, we proceed to export it to a CSV file:

```{r}
df |> write.csv("cleaned_data_R.csv", row.names = FALSE)
```
